VORTEX: Vision Transformers for Interpretable Temporal Dating of Historical Paintings Through Ordinal Classification

### 2025-06-09 report/PaperForReview.tex
updated paper

The precise temporal placement of artworks represents a fundamental challenge in art history scholarship \cite{Mishory00}. Accurate dating influences virtually every aspect of research, from understanding individual artistic development to tracing the evolution of broader cultural movements \cite{Elgammal18}. When researchers can confidently assign specific creation years to paintings, they unlock insights into how artistic innovations emerged, spread across geographic boundaries, and influenced subsequent generations of artists.

Despite centuries of scholarly effort and recent museum digitization initiatives, a substantial portion of the world's artistic heritage remains imprecisely dated. Major cultural institutions continue to struggle with temporal attribution, often cataloging significant works with vague temporal ranges such as ``first half of the 19th century'' rather than specific years.\footnote{See, for example, \url{https://collections.louvre.fr/en/ark:/53355/cl010064491}.} This imprecision creates cascading challenges for digital humanities research, hampering quantitative analyses of stylistic evolution and limiting our ability to understand chronological relationships between artworks.

@@

Traditional dating methods each have inherent limitations. Connoisseurship remains subjective and often contentious among experts. Documentary evidence frequently proves incomplete or ambiguous. Scientific analysis techniques require significant resources and often yield broad date ranges. Moreover, curators must invest substantial manual effort to narrow broad time ranges into single years—a process that remains labor-intensive and difficult to scale.

While computational methods offer potential to augment traditional approaches, previous efforts have primarily focused on related problems such as style classification and artist attribution, not directly addressing precise temporal dating with the interpretability necessary for scholarly acceptance.

@@

We introduce VORTEX (Vision Ordinal Regression Temporal EXtraction) to address the unique challenges of exact year prediction for historical paintings. Our approach targets Western European paintings from 1600 to 1899, encompassing major artistic movements from Baroque through early Impressionism. We formulate three core objectives:

@@

    \item \textbf{Interpretable decision-making}: Implement attention-based visualization techniques that reveal which visual elements inform temporal predictions, providing art historians with evaluable evidence.
\end{enumerate}

@@

\subsection{Expected Impact}

From a technical perspective, we demonstrate that fine-tuning smaller open-weight pre-trained models can rival the performance of large SOTA multimodal language models for domain-specific computer vision tasks. This suggests that specialized architectures with efficient adaptation methods can achieve competitive results without requiring massive computational resources.

From an art historical perspective, our emphasis on interpretability through attention analysis provides a crucial bridge between computational predictions and scholarly practices. By visualizing which visual elements inform temporal decisions, we enable art historians to evaluate model predictions against their domain expertise, building trust in computational methods within the humanities.

@@

The recent emergence of Vision Transformers represents the latest evolution. These architectures offer advantages through global self-attention mechanisms that capture long-range dependencies in visual data. DINOv2 models \cite{Oquab23}, trained through self-supervised learning on massive visual datasets, have demonstrated exceptional transfer capabilities suitable for specialized domains with scarce annotated examples.

@@

\begin{figure*}[t]
    \centering
    \begin{subfigure}[t]{\textwidth}
        \centering
        \includegraphics[width=\linewidth]{images/Distribution of Western Paintings in all Datasets by Year (1600-1899) all decades big.png}
        \caption{Distribution of Western Paintings Collected Across all Datasets by Decade (1600-1900)}
        \label{fig:subfig-a}
    \end{subfigure}

@@

To construct our dataset, we carefully evaluated prior work to identify sources that meet our research requirements. We aggregated paintings from four major collections, as shown in Table \ref{tab:dataset_comparison}.

\begin{enumerate}[leftmargin=1.2em,label=(\arabic*)]
    \item The Joconde database provides metadata for French museum artworks \cite{JocondeTerms}.
    \item WikiArt offers a user-curated collection across multiple movements \cite{WikiArtTerms}. 
    \item The Web Gallery of Art focuses on European paintings from Renaissance through 19th century \cite{WGATerms}.
    \item The Rijksmuseum collection contains Dutch and Flemish paintings \cite{Mensink14}.
\end{enumerate}

\begin{table*}[ht]
\centering
\caption{Dataset comparison across relevant features for year prediction of artworks.}
\label{tab:dataset_comparison}
\begin{tabular}{lccccc}
\toprule
\textbf{Feature \ Dataset} & \textbf{Wiki Art} & \textbf{WGA} & \textbf{Joconde} & \textbf{Rijks} & \textbf{\underline{All Datasets}}\\
\midrule
Number of Artworks           &       28997        &         10000      &       11954        &    1506     & 50781      \\
Publicly Available           &     \checkmark          &          \checkmark     &        \checkmark       &      \checkmark  & \checkmark       \\
Year Label Granularity $^{\dagger}$ &       0.2861        &       0.0096        &     0.0          &       0.0    & 0.1600    \\
Earliest Year                &        1600       &      1600         &      1600         &      1600      &  1600  \\
Latest Year                  &     1899          &        1899       &      1899         &       1899  & 1899      \\
Geographic Scope             &      \makecell{Western \\ Europe/America}         &       \makecell{Western \\ Europe/America}        &     \makecell{Predominantly \\ France}          &        Western Europe    & \makecell{Western \\ Europe/America} \\
\bottomrule
\end{tabular}
\noindent $^{\dagger}$ Average year 'window' range
\end{table*}

@@

\paragraph{Filtering} Our dataset artwork inclusion criteria impose the following constraints:
\begin{itemize}
    \item Works must be explicitly identified as paintings (encompassing oil, tempera, watercolor, and mixed media techniques)
    \item Works must originate from Western European contexts
    \item Each work must include a digitized image
    \item Each work must possess either an exact year annotation or a date range with uncertainty less than 10 years
\end{itemize}
We exclude drawings, prints, sculptures, and decorative arts to maintain focus on painted works where temporal stylistic evolution follows consistent patterns.

@@

VORTEX also achieved a perfect prediction on an 1899 portrait (Figure~\ref{fig:predictions}). The attention map reveals its strategy: it focused intensely on high-information regions, particularly the subject's stiff white cuffs and head. The model learned to correlate the painting style and fashion of the era with the late 19th century. This painting's visual cues were all stylistically consistent with its creation year, allowing the model to make a precise attribution.

@@

% --- Best Predictions Row ---
(a) \textbf{Best Predictions} \\[0.5em]
\begin{tabular}{cc}
    \includegraphics[width=0.45\linewidth]{images/best_vortex.jpg} &
    \includegraphics[width=0.4\linewidth]{images/best_gemini.jpg} \\
    \makecell{\textbf{VORTEX}\\ (0 year error)} & \makecell{\textbf{Gemini Flash 2.0}\\ (0 year error)} \\
\end{tabular}

@@

% --- Worst Predictions Row ---
(b) \textbf{Worst Predictions} \\[0.5em]
\begin{tabular}{cc}
    \includegraphics[width=0.5\linewidth]{images/worst_vortex.jpg} &
    \includegraphics[width=0.35\linewidth]{images/worst_gemini.jpg} \\
    \makecell{\textbf{VORTEX}\\ (290 yrs error)} & \makecell{\textbf{Gemini Flash 2.0} \\ (441 yrs error)} \\
\end{tabular}

@@

The performance and failure modes of both VORTEX and Gemini 2.0 Flash demonstrate that these AI tools cannot replace expert scholarship but serve as powerful supplementary instruments. For curatorial applications, VORTEX provides immediate practical value by rapidly suggesting temporal attributions for new acquisitions and offering independent evidence for disputed dates. The attention visualizations provide interpretable evidence that curators can evaluate against their expertise, building trust through transparency.

@@

{\small
\bibliographystyle{ieee_fullname}
\bibliography{egbib}
}
\end{document}

>>>>

The precise temporal placement of artworks represents a fundamental challenge in art history scholarship \cite{Mishory00}. Accurate dating influences virtually every aspect of research, from understanding individual artistic development to tracing the evolution of broader cultural movements \cite{Elgammal18}. 
% When researchers can confidently assign specific creation years to paintings, they unlock insights into how artistic innovations emerged, spread across geographic boundaries, and influenced subsequent generations of artists.

Despite centuries of scholarly effort and recent museum digitization initiatives, a substantial portion of the world's artistic heritage remains imprecisely dated. Major cultural institutions continue to struggle with temporal attribution, often cataloging significant works with vague temporal ranges such as ``first half of the 19th century'' rather than specific years.\footnote{See, for example, \url{https://collections.louvre.fr/en/ark:/53355/cl010064491}.} This imprecision creates challenges for digital humanities research, limiting our ability to understand chronological relationships between artworks.

@@

Traditional dating methods each have inherent limitations. Connoisseurship remains subjective and often contentious among experts. Scientific analysis techniques require significant resources and often yield broad date ranges. Moreover, curators must invest substantial manual effort to narrow broad time ranges into single years—a process that is labor-intensive and difficult to scale.

While computational methods offer potential to augment traditional approaches, previous efforts have primarily focused on related problems such as style classification and artist attribution, not directly addressing precise temporal dating.
%with the interpretability necessary for scholarly acceptance.

@@

We introduce \textbf{VORTEX} (\textbf{V}ision \textbf{O}rdinal \textbf{R}egression \textbf{T}emporal \textbf{EX}traction) to address the unique challenges of exact year prediction for historical paintings. Our approach targets Western European paintings from 1600 to 1899, encompassing major artistic movements from Baroque through early Impressionism. We formulate two core objectives:

@@

    \item \textbf{Interpretable decision-making}: Implement attention-based visualization techniques that reveal which visual elements inform temporal predictions, providing art historians with understandable evidence.
\end{enumerate}

@@

% \subsection{Expected Impact}

% From a technical perspective, we demonstrate that fine-tuning smaller open-weight pre-trained models can rival the performance of large SOTA multimodal language models for domain-specific computer vision tasks. This suggests that specialized architectures with efficient adaptation methods can achieve competitive results without requiring massive computational resources.

% From an art historical perspective, our emphasis on interpretability through attention analysis provides a crucial bridge between computational predictions and scholarly practices. By visualizing which visual elements inform temporal decisions, we enable art historians to evaluate model predictions against their domain expertise, building trust in computational methods within the humanities.

@@

The recent emergence of Vision Transformers offer advantages through global self-attention mechanisms that capture long-range dependencies in visual data. DINOv2 models \cite{Oquab23}, trained through self-supervised learning on massive visual datasets, have demonstrated exceptional transfer capabilities suitable for specialized domains with scarce annotated examples.

@@

@@

To construct our dataset, we carefully evaluated prior work to identify sources that meet our research requirements. We aggregated paintings from four major collections, as shown in Table \ref{tab:dataset_comparison}. Appendix \ref{fig:subfig-a} contains a further granular breakdown of the dataset.

% \begin{enumerate}[leftmargin=1.2em,label=(\arabic*)]
%     \item The Joconde database provides metadata for French museum artworks \cite{JocondeTerms}.
%     \item WikiArt offers a user-curated collection across multiple movements \cite{WikiArtTerms}. 
%     \item The Web Gallery of Art focuses on European paintings from Renaissance through 19th century \cite{WGATerms}.
%     \item The Rijksmuseum collection contains Dutch and Flemish paintings \cite{Mensink14}.
% \end{enumerate}

\begin{table*}[ht]
\centering
\caption{Dataset comparison across relevant features for year prediction of artworks.}
\label{tab:dataset_comparison}
\begin{tabular}{lccccc}
\toprule
\textbf{Feature \ Dataset} & \textbf{Wiki Art}\cite{WikiArtTerms} & \textbf{WGA}\cite{WGATerms} & \textbf{Joconde}\cite{JocondeTerms} & \textbf{Rijks}\cite{Mensink14} & \textbf{\underline{All Datasets}}\\
\midrule
Number of Artworks           &       28997        &         10000      &       11954        &    1506     & 50781      \\
Publicly Available           &     \checkmark          &          \checkmark     &        \checkmark       &      \checkmark  & \checkmark       \\
Year Label Granularity $^{\dagger}$ &       0.2861        &       0.0096        &     0.0          &       0.0    & 0.1600    \\
Earliest Year                &        1600       &      1600         &      1600         &      1600      &  1600  \\
Latest Year                  &     1899          &        1899       &      1899         &       1899  & 1899      \\
Geographic Scope             &      \makecell{Western \\ Europe/America}         &       \makecell{Western \\ Europe/America}        &     \makecell{Predominantly \\ France}          &        Western Europe    & \makecell{Western \\ Europe/America} \\
\bottomrule
\end{tabular}
\noindent $^{\dagger}$ Average year 'window' range
\end{table*}

@@

\paragraph{Filtering} Our dataset artwork inclusion criteria impose the following constraints:
\begin{itemize}
    \item Works must be explicitly identified as paintings (encompassing oil, tempera, watercolor, and mixed media techniques)
    \item Works must originate from Western European contexts
    \item Each work must include a digitized image
    \item Each work must possess either an exact year annotation or a date range with uncertainty less than 10 years
\end{itemize}
We exclude drawings, prints, sculptures, and decorative arts to maintain focus on painted works where temporal stylistic evolution follows consistent patterns. See Appendix \ref{fig:subfig-e} for example images from the dataset.

@@

VORTEX also achieved a perfect prediction on an 1889 portrait (Figure~\ref{fig:predictions}). The attention map reveals its strategy: it focused intensely on high-information regions, particularly the subject's stiff white cuffs and head. The model learned to correlate the painting style and fashion of the era with the late 19th century. This painting's visual cues were all stylistically consistent with its creation year, allowing the model to make a precise attribution.

@@

% --- Best Predictions Row ---
(a) \textbf{Best Predictions} \\[0.5em]
\begin{tabular}{cc}
    \includegraphics[width=0.45\linewidth]{images/best_vortex.jpg} &
    \includegraphics[width=0.4\linewidth]{images/best_gemini.jpg} \\
    \makecell{\textbf{VORTEX}\\ (0 year error) \\ Ground Truth: 1889} & \makecell{\textbf{Gemini Flash 2.0}\\ (0 year error) \\ Ground Truth: 1892} \\
\end{tabular}

@@

% --- Worst Predictions Row ---
(b) \textbf{Worst Predictions} \\[0.5em]
\begin{tabular}{cc}
    \includegraphics[width=0.5\linewidth]{images/worst_vortex.jpg} &
    \includegraphics[width=0.35\linewidth]{images/worst_gemini.jpg} \\
    \makecell{\textbf{VORTEX}\\ (290 yrs error)\\ Ground Truth: 1898} & \makecell{\textbf{Gemini Flash 2.0} \\ (441 yrs error)\\ Ground Truth: 1861} \\
\end{tabular}

@@

The performance and failure modes of both VORTEX and Gemini 2.0 Flash demonstrate that these AI tools cannot replace expert scholarship but serve as powerful supplementary instruments. 
% For curatorial applications, VORTEX provides immediate practical value by rapidly suggesting temporal attributions for new acquisitions and offering independent evidence for disputed dates.
The attention visualizations provide interpretable evidence that curators can evaluate against their expertise, building trust through transparency.

@@

{\small
\bibliographystyle{ieee_fullname}
\bibliography{egbib}
}

\newpage

\begin{figure*}[t]
\section*{Appendix}
    \centering
    \begin{subfigure}[t]{\textwidth}
        \centering
        \includegraphics[width=\linewidth]{images/Distribution of Western Paintings in all Datasets by Year (1600-1899) all decades big.png}
        \caption{Distribution of Western Paintings Collected Across all Datasets by Decade (1600-1900)}
        \label{fig:subfig-a}
    \end{subfigure}

\end{document}
